On the Synchronization Power of the Openflow Data Plane

### 2015-05-14 inband-sync.tex
all petr's comments are either fixed or added to the tex

\newcommand{\stefan}[1]{\textit{\textcolor{red}{[stefan]: #1}}} % Comments
\newcommand{\liron}[1]{\textit{\textcolor{mygreen}{[liron]: #1}}} % Comments
\newcommand{\petr}[1]{\textit{\textcolor{blue}{[carlos]: #1}}} % Comments

@@

\section{Introduction}\label{sec:intro}

By consolidating and outsourcing the control over the dataplane switches to a logically
centralized controller, Software-Defined Networks (SDNs)
simplify network management and facilitate faster innovations:
the (software) control plane can evolve independently from the
(hardware) data plane.
Moreover, Openflow, the standard SDN protocol today, introduces many generalizations,
in terms of traffic engineering, definition of flows, as well as in-band network functionalities,
by relying on a simple match-action paradigm which allows us to define
forwarding rules based not only on Layer-2, but also Layer-3 and Layer-4 (and possibly beyond).

@@

By providing more complex and atomic transactions over the
switch configuration
space, several synchronization problems can be resolved.
This paper investigates the implementation of a transactional interface
(\`{a} la STN~\cite{stn}) modified concurrently by multiple controllers.

@@

\section{Background}\label{sec:background}

The transactional interface presented in this paper can be implemented in
Openflow~\cite{of-spec}, the standard SDN protocol today.
% In this section,
%we provide the necessary background.
In a nutshell,
Openflow
follows a match-action concept: Openflow switches store
rules (installed by the controller) consisting of a match and an
action part. A packet matched by a certain rule will be subject
to the associated action (to be immediately applied now or later).
For example, an action can define a port to which the
matched packet should be forwarded, or add or change a tag
(a certain part in the packet header).
Each Openflow switch stores one or multiple flow tables,
each of which contains a set of rules: the \emph{flow entries}.

Flow
tables form a pipeline, and flow entries are ordered according
to priorities: A packet arriving at a switch is first checked by
the rule of the highest priority in table 0: the fields of the
data packet are compared with the match fields of that rule,
and if they fit, some instructions (the actions) are executed;
subsequently, lower priority rules are checked. Depending on
the outcome of the table 0 processing, the packet may be sent
to additional flow tables in the pipeline.
Instructions can be
used to define additional tables to be visited (a ``goto instruction''),
to modify the set of actions to be applied later (either by appending,
deleting, or modifying actions), or to immediately apply some
actions to the packet.

@@

\paragraph{The FlowMod Command}
The main messages used in Openflow
to update the switch configuration is called \emph{FlowMod}.
In a nutshell, \emph{FlowMod} either specifies a new flow entry or
modifications to existing flow entries.
A flow entry consists of: (1) a match defined by a ternary pattern over packet header fields
(a value and a mask), and (2) a set of actions.
%,a match and an action.
The standard behavior is that the new flow entry will simply \emph{replace}
an existing one if the existing one has exactly the \emph{same match},
but will be installed
\emph{in addition} to existing entries if they only
partially overlap (or define flowspace subsets or supersets).
The latter can result
in inconsistencies, especially if the old and the new rule
have the same priority.

@@

\paragraph{Bundling}
Another interesting feature in Openflow
is the \emph{bundle}. A bundle
aggregates multiple control commands (e.g., \emph{FlowMod} messages)
into one message, with \emph{all-or-nothing semantics}:
either all of the commands are
performed, or none of them is. In particular, if, for some reason, the
configuration request contained in one of the bundled commands cannot
be performed, all the other requests  are rejected and an error
message is sent to the controller that issued them.
When properly configured,
the bundled processing can be done in a
\emph{packet-atomic} way, in the sense that
each packet is processed according
to the switch configuration either \emph{before} or \emph{after} the
bundle (i.e., all its commands) took effect.
If multiple \emph{FlowMod} messages with the $\checko$ flag are sent as a single bundle,
the bundle command also returns a reason
why the overlap checks failed: it returns the failed command in the bundle.
Concretely, when the commit fails, the switch must generate an error message corresponding to the
message that could not be applied. The error message must use the identifier (\emph{xid}) of the offending message,
the error data field corresponding to that message and the error code corresponding to the failure. If
multiple messages are generating errors, the switch may return only the first error found or generate
multiple error messages for the same bundle.

@@

In this paper, we will use the notation
%$\textbf{bundle}\{\;\ldots\;\}\rightarrow res$ to refer to represent a
$\textbf{bundle}\{\;\ldots\;\}$ to refer to a
sequence of commands sent in a bundle. The bundle returns
$\top$ if and only if
all commands are performed successfully, and otherwise $\bot$.
The bundles used in this paper will always have the following structure:
\begin{enumerate}
\item First, it contains a \emph{create bundle} message: a \texttt{ofpt\_bundle\_control} message of type
\texttt{ofpbct\_open\_request}.
\item The follows a sequence of commands (e.g., \emph{FlowMod}), each wrapped in an
\emph{add to bundle message} (\texttt{ofpt\_bundle\_add\_message}) with the \emph{bundle identifier}.
\item Finally, a \emph{commit bundle} control message (\texttt{ofpt\_bundle\_control}) of
type \texttt{ofpbct\_commit\_request}.
\end{enumerate}
Again, the bundle always returns a binary result: \emph{error} ($\bot$) or \emph{success} ($\top$).

@@

\paragraph{Remarks}
As we will see, in our implementation,
the switch \emph{configuration} consists of two parts:
the ``normal'' configuration which is defined by the network policy and
directly effects the traffic (e.g., forwarding rules), and some additional
configuration which serves as synchronization information 
for our mechanism. 
By installing Openflow \emph{flow monitors}, 
a controller can read the entire configuration 
(and subscribe for updates). 
%
%can i match multiple flows in multiple tables? yes, but only for FlowMod, not GroupMod: but that is good enough!%

@@

At first sight, the Openflow bundle feature seems to provide everything
we need to implement our transactional interface: the bundle
 defines a sequence of read and write operations
 on the switch state which is executed atomically.
 However, while the bundle can indeed be seen
 as a ``mini-transaction'', it comes with several
 limitations, which we discuss in the following.

First, if employed without any additional features, a bundle does not allow
 to change the switch state in any non-trivial way:
the read and write actions in the bundle do
 not support any conditional \emph{modifications} or computations which
 depend on the read results.
But also using the bundle together with the check overlap feature ($\checko$)
comes with limitations: while it is possible
 to \emph{abort} transactions depending on certain conditions
 in the \emph{overlapping
 flow spaces}, oftentimes it is desirable to be able
 to define more general conditions and notions of conflicts
 under which transactions should not take effect. Moreover,
 sometimes it is desirable to react to conflicts in smarter ways than
 to simply abort, e.g.,
 by supporting \emph{modifications}.
Let us consider two examples.

@@

\item \textbf{Example 2: Supporting modifications.} Consider the problem of modifying previously
installed forwarding policies.
While a bundle allows to conditionally add new rules or remove
old rules (and, using the flow space overlap
check feature $\checko$, to abort the transaction if potentially conflicting
rules were already there), it cannot be used
to modify rules in non-trivial ways.
In other words, the bundle and overlap check only help
for the \emph{conditional creation} of rules, not their \emph{modification}.
\end{enumerate}

@@

Our approach is based on the distributed management of \emph{identifiers (IDs)}. In particular,
we introduce two new synchronization objects (and describe their implementation in standard Openflow),
which allow controllers to coordinate
themselves based on these identifiers.

As we will see, in the Openflow implementation of our mechanism, 
in addition to the ``normal'' configuration stored at the switch
(e.g., forwarding rules according to network policies), 
controllers will maintain additional information 
in the switch configuration: this information is used by the
controllers to synchronize.

\subsection{Synchronization Objects}\label{sec:t-if}

@@

As we will see, the bundle feature is not required to implement these semantics.
Also note that memory read operations correspond to reading the switch
configuration, and are simple as they do not affect the execution of the transaction
(unlike the $\textit{\compare}$ operation).

@@

@@

@@

\paragraph{The resource claimer}
%features , including
%\emph{bundling}, \emph{conditional application}, and \emph{groups}.
%[[PK let us leave for the moment
%We will describe later how to implement some of these features
%in older Openflow versions where they are not
%supported. \textbf{TODO}
%]]
Algorithms~\ref{alg:claim}-\ref{alg:cas} describe implementations of
transactional operations  $\textit{claim}(x)$, $\textit{\claimcheck}(x)$,
$\textit{unclaim}(x)$, $\textit{write}(x)$,  $\textit{\compare}(x)$
and $\textit{\cas}(x)$, respectively.
Algorithm~\ref{alg:commit} describes the procedure executed by
a transaction at \emph{commit time}.

We describe each operation as a message factory that returns OpenFlow commands which
should be sent to the switch as part of the same transaction.
We assume that every transaction has a unique identifier and is used to start a \emph{bundle}.
All messages generated during a transaction invocation, are wrapped by bundle messages carrying
the bundle's ID and followed by bundle commit message.
Messages with same bundle ID are considered to belong to same bundle and are performed atomically.

@@

Our implementation is based on the \emph{FlowMod} command.
We use the $\checko$ optional flag to make the transaction execution dependent on switch state.
We consider the match part of an entry to be a ternary string of bounded length (represented by binary value and mask)
which can be applied to any set of supported packet header fields. In our implementation, the $64$bit meta-data field alone is
already sufficient. Moreover we consider the action part of an entry to be an integer, which in practice can be implemented by a
\emph{set meta-data} instruction where the written value is that integer.

@@

\subsection{CAS and Consensus}

Given our memory interface and using the bundle concept, we can implement
a powerful \emph{Compare-And-Set} (CAS) operation which takes two integer parameters
$\textit{old}$ and $\textit{new}$, reads the register and, if the
value of it is $\textit{old}$, replaces it with $\textit{new}$ and
returns $\top$; otherwise, $\bot$ is returned and
we say that the CAS operation \emph{fails} (which causes an abort of
the invoking transaction). As can be seen Algorithm \ref{alg:cas}, CAS is implemented by two commands. These commands should be sent as part of the same bundle

@@

@@

>>>>

\newcommand{\stefan}[1]{\textit{\textcolor{red}{[stefan]: #1}}} % Comments
\newcommand{\liron}[1]{\textit{\textcolor{mygreen}{[liron]: #1}}} % Comments
\newcommand{\petr}[1]{\textit{\textcolor{blue}{[petr]: #1}}} % Comments

@@

\section{Introduction}\label{sec:intro}
By consolidating and outsourcing the control over the dataplane switches to a logically
centralized controller, Software-Defined Networks (SDNs)
simplify network management and facilitate faster innovations:
the (software) control plane can evolve independently from the
(hardware) data plane.
Moreover, Openflow, the standard SDN protocol today, introduces many generalizations,
in terms of traffic engineering, definition of flows, as well as in-band network functionalities,
by relying on a simple match-action paradigm which allows us to define
forwarding rules based not only on Layer-2, but also Layer-3 and Layer-4 (and possibly beyond).

@@

By providing more complex and atomic transactions over the
switch configuration
space, several synchronization problems can be resolved.
This paper investigates the implementation of a transactional interface
(\`{a} la STN~\cite{stn}) modified concurrently by multiple controllers.
\petr{"interface modified concurrently" does not sound right -> "accessed concurrently"?}

@@

\section{Background}\label{sec:background}
\petr{ the 1st sentence does not belong here}\liron{I think it serves as a motivation for the section, but it fines by me to remove it}
The transactional interface presented in this paper can be implemented in
Openflow~\cite{of-spec}, the standard SDN protocol today.
% In this section,
%we provide the necessary background.
In a nutshell,
Openflow
follows a match-action concept: Openflow switches store
rules (installed by the controller) consisting of a match and an
action part. A packet matched by a certain rule will be subject
to the associated action (to be immediately applied now or later).\petr{"to be immediately applied now or later" - sounds funny}\liron{I agree it sounds a bit strange, maybe we should explain this later after we disscuss the pipeline}
For example, an action can define a port to which the
matched packet should be forwarded, or add or change a tag
(a certain part in the packet header).
Each Openflow switch stores one or multiple flow tables,
each of which contains a set of rules: the \emph{flow entries}.

Flow
tables form a pipeline, and flow entries are ordered according
to priorities: A packet arriving at a switch is first checked by
the rule of the highest priority in table 0: the fields of the
data packet are compared with the match fields of that rule,
and if they fit, some instructions (the actions)  are executed; \petr{ what is the difference between instruction or action?} \liron{we should explain that: every packet is associated with an action set. This set is empty at the beginning and is changed according to the flow table matches. Each match is asociated with instructions that can add or modify actions in the action set and also to apply (now) all actions in the action set. By default action set is applied at the end of the pipeline.}
subsequently, lower priority rules are checked. Depending on
the outcome of the table 0 processing, the packet may be sent
to additional flow tables in the pipeline.
Instructions can be
used to define additional tables to be visited (a ``goto instruction''),
to modify the set of actions to be applied later (either by appending,
deleting, or modifying actions), or to immediately apply some
actions to the packet.

@@

\paragraph{The FlowMod Command}
The main messages used in Openflow
to update the switch configuration is called \emph{FlowMod}.
In a nutshell, \emph{FlowMod} either specifies a new flow entry or
modifications to existing flow entries.
A flow entry consists of: (1) a match defined by a ternary pattern over packet header fields
(a value and a mask), (2) a set of actions, and (3) a priority.
%,a match and an action.
The standard behavior is that the new flow entry will simply \emph{replace}
an existing one if the existing one has exactly the \emph{same match},
but will be installed
\emph{in addition} to existing entries if they only
partially overlap (or define flowspace subsets or supersets).
The latter can result
in inconsistencies, especially if the old and the new rule
have the same priority.

@@

\paragraph{Bundling}

Another interesting feature in Openflow
is the \emph{bundle}. A bundle allows to send multiple commands (e.g., \emph{FlowMod} messages) with the same bundle id and then execute them  by sending a bundle commit message. Theses (bundled) commands are performed with \emph{all-or-nothing semantics}:
either all of the commands are
performed, or none of them is. In particular, if, for some reason, the
configuration request contained in one of the bundled commands cannot
be performed, all the other requests  are rejected and an error
message is sent to the controller that issued them.
When properly configured,
the bundled processing can be done in a
\emph{packet-atomic} way, in the sense that
each packet is processed according
to the switch configuration either \emph{before} or \emph{after} the
bundle (i.e., all its commands) took effect. \liron{we need to disscuss controller atomicity and order preservation inside a bundle}
If multiple \emph{FlowMod} messages with the $\checko$ flag are sent as a single bundle,
the bundle command also returns a reason
why the overlap checks failed: it returns the failed command in the bundle.
Concretely, when the commit fails, the switch must generate an error message corresponding to the
message that could not be applied. The error message must use the identifier (\emph{xid}) of the offending message,
the error data field corresponding to that message and the error code corresponding to the failure. If
multiple messages are generating errors, the switch may return only the first error found or generate
multiple error messages for the same bundle.

@@

In this paper, we will use the notation
%$\textbf{bundle}\{\;\ldots\;\}\rightarrow res$ to refer to represent a
$\textbf{bundle}\{\;\ldots\;\}$ to refer to a
sequence of commands sent in a bundle. The bundle returns
$\top$ if and only if
all commands are performed successfully, and otherwise $\bot$.
\petr{What about replacing $\top-\bot$ with ack-nack? Can we also get a reason why a bundle did not commit?}
The bundles used in this paper will always have the following structure:
\begin{enumerate}
\item First, it contains a \emph{create bundle} message: a \texttt{ofpt\_bundle\_control} message of type
\texttt{ofpbct\_open\_request}.
\item The follows a sequence of commands (e.g., \emph{FlowMod}), each wrapped in an
\emph{add to bundle message} (\texttt{ofpt\_bundle\_add\_message}) with the \emph{bundle identifier}.
\item Finally, a \emph{commit bundle} control message (\texttt{ofpt\_bundle\_control}) of
type \texttt{ofpbct\_commit\_request}.
\end{enumerate}
Again, the bundle always returns a binary result: \emph{error} ($\bot$) or \emph{success} ($\top$).

@@

\paragraph{Remarks}
As we will see, in our implementation,
the switch \emph{configuration} consists of two parts:
the ``normal'' configuration which is defined by the network policy and
directly effects the traffic (e.g., forwarding rules), and some additional
configuration which serves as synchronization information 
for our mechanism. 
By installing Openflow \emph{flow monitors} with the \textsf{OFPFMF\_INITIAL} optional flag,  
a controller can read the current configuration. 
% %(and subscribe for updates). 
%
%can i match multiple flows in multiple tables? yes, but only for FlowMod, not GroupMod: but that is good enough!%
\liron{I changed a bit the description of flow monitors following Petr's comment, I hope it's more clear now.}

@@

 generic way for multiple distributed controllers to safely update switch configuration with no need of any control plane

At first sight, the Openflow bundle feature seems to provide everything
we need to implement our transactional interface \petr{did we say what we need already?}: the bundle
 defines a sequence of read and write operations
 on the switch state which is executed atomically.
 However, while the bundle can indeed be seen
 as a ``mini-transaction'', it comes with several
 limitations, which we discuss in the following.
 \petr{unclear what "reads and writes" are here, not discussed before. How openflow can be used to implement reads and writes?}

\liron{following Petr's comments, I suggest the following paragraph}
At first sight, the Openflow bundle feature seems to provide a simple and generic way for a controller to atomically update switch configuration, similarly to a transaction in shared memory system.
However, while the bundle can indeed be seen
 as a ``mini-transaction'', it comes with several
 limitations, which we discuss in the following.

\liron{ 
First, if employed without any additional features, a bundle does not allow
 to change the switch state in any non-trivial way:
the commands in the bundle essentially do
 not support any conditional \emph{modifications} or computations which depend on any current switch configuration (except to limited types of constraints, e.g., free space).
 }

But also bundling FlowMod commands with the check overlap feature ($\checko$)
comes with limitations: while it is possible
 to \emph{abort} a bundle depending on certain conditions
 in the \emph{overlapping
 flow spaces}, oftentimes it is desirable to be able
 to define more general conditions and notions of conflicts
 under which an atomic update should not take effect.
 \liron{replaced transaction in the last sentence with " atomic update"}

  Moreover,
 sometimes it is desirable to react to conflicts in smarter ways than
 to simply abort, e.g.,
 by supporting \emph{modifications}.
Let us consider two examples.

@@

\item \textbf{Example 2: Supporting modifications.} Consider the problem of modifying previously
installed forwarding rules.
While a bundle allows to conditionally add new rules or remove
old rules (and, using the flow space overlap
check feature $\checko$, to abort the transaction if potentially conflicting
rules were already there), it cannot be easily used
to conditionally modify existing rules .
In other words, the bundle and overlap check only help
for the \emph{conditional creation} of rules, not their \emph{modification}.
\end{enumerate}

@@

Our approach is based on the distributed management of \emph{identifiers (IDs)} \petr{unclear what "identifiers" mean here}. In particular,
we introduce two new synchronization objects (and describe their implementation in standard Openflow),
which allow controllers to coordinate
themselves based on these identifiers.

As we will see, in the Openflow implementation of our mechanism, 
in addition to the ``normal'' configuration stored at the switch
(e.g., forwarding rules according to network policies), 
controllers will maintain additional information 
in the switch configuration: this information is used by the
controllers to synchronize. 
\petr{"As we will see,..." - looks funny compared with "As we will see ..." at the end of Sec 2}
\subsection{Synchronization Objects}\label{sec:t-if}

@@

\petr{it is unclear why exactly we need these write and compare operations}

As we will see, the bundle feature is not required to implement these semantics.
Also note that memory read operations correspond to reading the switch
configuration, and are simple as they do not affect the execution of the transaction
(unlike the $\textit{\compare}$ operation). \petr{read operations "do not affect the execution of the transaction" appears strange, given that a transaction, if committed, "appears to execute sequentially". Something seems wrong here...}

@@

\petr{Would be great actually to motivate why we use reads, writes and claim ids. Maybe skecth a solution to some important problems using these primitives in the transactions.} \liron{we mention that its kind of locking. but I agree we can motivate it better}

@@

\petr{what is "offset" here (offset addr may correspond")?}\liron{I don't understand the comment}

@@

%features , including
%\emph{bundling}, \emph{conditional application}, and \emph{groups}.
%[[PK let us leave for the moment
%We will describe later how to implement some of these features
%in older Openflow versions where they are not
%supported. \textbf{TODO}
%]]
Algorithms~\ref{alg:claim}-\ref{alg:compare} describe implementations of
transactional operations  $\textit{claim}(x)$, $\textit{\claimcheck}(x)$,
$\textit{unclaim}(x)$, $\textit{write}(x)$ and $\textit{\compare}(x)$, respectively.
%Algorithm~\ref{alg:commit} describes the procedure executed by
%a transaction at \emph{commit time}.

We describe each operation as a message factory \petr{did not get what is "message factory"} that returns OpenFlow commands which
should be sent to the switch as part of the same transaction.
We assume that every transaction has a unique identifier and is used to start a \emph{bundle}.
All messages generated during a transaction invocation, are wrapped by bundle messages carrying
the bundle's ID and followed by bundle commit message.
Messages with same bundle ID are considered to belong to same bundle and are performed atomically.
\liron{I don't think we need all of this bundle staff here. we need to move it to next sextion - policy updates}

@@

Our implementation is based on the \emph{FlowMod} command.
We use the $\checko$ optional flag to make the transaction execution dependent on switch state \liron{ last sentence might a bit contradict what we said in the limitation section}.
We consider the match part of an entry to be a ternary string of bounded length (represented by binary value and mask)
which can be applied to any set of supported packet header fields. In our implementation, the $64$bit meta-data field alone is
already sufficient. Moreover we consider the action part of an entry to be an integer, which in practice can be implemented by a
\emph{set meta-data} instruction where the written value is that integer.

\paragraph{The resource claimer}

\petr{
the "resource claimed" part a hard read, the trick of using masks and values is smart, deserves to be explained better
}

\petr{Alg 2 returns "cmd1, cmd2", what does it mean? Isn't it supposed to return a boolean?}
\petr{what is this "$\concat$" operation? simple concatenation?}
\petr{sometimes FlowMod commands have flag and action, sometimes not, sometimes flag=0, semantics of all this is unclear.}

@@

\subsection{CAS and Consensus}
\petr{ some motivation of why CAS is important/useful is missing}
Given our memory interface and using the bundle concept, we can implement
a powerful \emph{Compare-And-Set} (CAS) operation which takes two integer parameters
$\textit{old}$ and $\textit{new}$, reads the register and, if the
value of it is $\textit{old}$, replaces it with $\textit{new}$ and
returns $\top$; otherwise, $\bot$ is returned and
we say that the CAS operation \emph{fails} (which causes an abort of
the invoking transaction). As can be seen Algorithm \ref{alg:cas}, CAS is implemented by two commands. These commands should be sent as part of the same bundle

@@

\petr{when talking about replicated state machines, we should make it clear that we that we talk about replication among controllers.}

\petr{"more complex operations": more complex than what?}

\petr{the discussion of CAS in STN is confusing, given that we never described STN.}

@@

\petr{this is the first time we talk concretely about "policies" ("forwarding policies"?), but they were mentioned a few times before. Shouldn't we have a short "SDN primer" in Sec 3?}

\petr{did not get what a "composition predicate" is}

\petr{why apply-update is "local" ("local update function")?}

\petr{the 1st paragraph is hard to digest}

\petr{Sec 5.2 relies too heavily on [3]. Do we expect the reader unfamiliar with [3] to get it?}
